Global code motion of congruent instructions

### 2017-04-17 paper.tex
abstract

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
We present a global code motion compiler optimization which schedules congruent
instructions across the program.  Not only this code motion saves code size, it
exposes redundancies in some cases, it exposes more instruction level
parallelism in the basic block to which instructions are moved, and it enables
other passes like loop invariant motion to remove more redundancies. The cost
model to drive the code motion is based on live range analysis on \SSA{}
representation such that the (virtual) register pressure does not increase.

We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
infrastructure available in \LLVM{}. The experimental results show an average of
2\% reduction in the number of spills, and saving of 1\% on overall compile time
on \SPEC{}. This being a code-size optimization, enables more inlining in
majority of the benchmarks. We have also seen execution
time improvements of approximately 5\% in a couple of \SPEC{} benchmarks
viz. \GCC{} and mcf, moreover, register spills reduced for almost all the
\SPEC{} bencmarks when compiled for X86\_64. This is an optimistic algorithm in
the sense that it considers all identical computations in a function as
potential candidates. We make an extra effort to hoist candidates by
partitioning the potential candidates in a way to enable partial hoisting in
case common hoisting points for all the candidates cannot be found. We also
formalize how register pressure will reduce as a result of code-motion and why
sorting the list of potential candidates w.r.t. their depth first numbers helps
hoist more candidates with less compile time overhead.
\end{abstract}

@@

@@

\begin{verbatim}
    b = m
    c = n
B0: if c is true then goto B3 else goto B4

@@

Barriers are based on the concept of pinned instructions \cite{click1995global}
but extended to adapt to \LLVM{} IR. Since a basic block in \LLVM{} IR is actually an
extended basic block because there might be calls in the middle of a basic block
which might not return. So barriers can be present in the middle of a basic
block. Essentially, any instruction that cannot guarantee progress is marked
as a barrier. In the absence of context (as in our current implementation), some
instructions which might be safe are still classified as barriers, e.g., volatile
loads/stores, calls with missing attributes.

\begin{verbatim}
computing downward-safety of hoistable instructions at hoist-point
Worklist = list of all basic blocks of hoistable instructions
for each basic block B in the dominator tree starting at hoist-point:
  if Worklist is empty
    return false // Path exists! not downward safe
  if B is in Worklist:
    remove B from worklist
    remove subtree with root at B // Available at B => downward safe from B
  if B is a leaf basic block:
    return false  // Path exists! not downward safe
  if B dominates hoist-point:
    return false  // Back edge

% Explain the overall algorithm, explain how anticipability is computed on demand
% explain the pinned-instructions/barriers.
\begin{verbatim}
Analyses available: Dominator Tree, DFS Numbering, Memory SSA,

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{|l|r|r|r|r|r|r|r|r|}
      \hline
  &  \multicolumn{2}{|c|}{Hoisted by LICM} &\multicolumn{2}{|c|}{Sunk by LICM} &\multicolumn{2}{|c|}{Functions Inlined} &  \multicolumn{2}{|c|}{Functions Deleted} \\\hline
 Benchmark        & trunk       & GCM	& trunk	& GCM	& trunk	& GCM	& trunk	& GCM  \\\hline
 400.perlbench	  & 2428	& 2480	& 36	& 35	& 1562	& 1565	& 232	& 234  \\\hline
 401.bzip2	  & 721	        & 825	& 3	& 3	& 172	& 180	& 32	& 33   \\\hline
 403.gcc	  & 8867	& 10079	& 143	& 182	& 5034	& 5080	& 1303	& 130  \\\hline
 429.mcf	  & 55	        & 51	& 1	& 1	& 4	& 4	& 0 	& 0    \\\hline
 433.milc	  & 2537	& 2795	& 3	& 3	& 75	& 81	& 0	& 0    \\\hline
 444.namd	  & 2204	& 2206	& 240	& 240	& 387	& 390	& 57	& 55   \\\hline
 445.gobmk	  & 3090	& 3107	& 18	& 19	& 735	& 775	& 230	& 234  \\\hline
 447.dealII	  & 59114	& 95884	& 60	& 8	& 100891& 100692& 30219	& 29966\\\hline
 450.soplex	  & 4668	& 4848	& 18	& 17	& 9801	& 9784	& 3125	& 3112 \\\hline
 453.povray	  & 9533	& 9079	& 23	& 32	& 5522	& 5532	& 1202	& 1201 \\\hline
 456.hmmer	  & 3200	& 3232	& 3	& 8	& 562	& 562	& 103	& 101  \\\hline
 458.sjeng	  & 400	        & 446	& 0	& 0	& 127	& 135	& 18	& 18   \\\hline
 462.libquantum	  & 362	        & 386	& 0 	& 0	& 87	& 87	& 21	& 21   \\\hline
 464.h264ref	  & 17075	& 17153	& 4	& 4	& 338	& 382	& 72	& 72   \\\hline
 470.lbm	  & 39	        & 39	& 0	& 0	& 13	& 13	& 4	& 4    \\\hline
 471.omnetpp	  & 565	        & 661	& 32	& 30	& 4790	& 4843	& 1321	& 1316 \\\hline
 473.astar	  & 691	        & 754	& 0	& 0	& 359	& 363	& 113	& 113  \\\hline
 482.sphinx3	  & 1599	& 1693	& 11	& 9	& 198	& 194	& 53	& 53   \\\hline
 483.xalancbmk	  & 9997	& 11453	& 56	& 54	& 69534	& 68839	& 29875	& 29519\\\hline
    \end{tabular}
  \end{center}
  \caption{Static metrics before and after code-hoisting on SPEC2006 at -Ofast}
  \label{tab:stats}
\end{table}

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
We present a global code motion (GCM) compiler optimization which schedules
congruent instructions across the program.  Not only this saves code size, it
exposes redundancies in some cases, it exposes more instruction level
parallelism in the basic block to which instructions are moved, and it enables
other passes like loop invariant motion to remove more redundancies. The cost
model to drive the code motion is based on live range analysis on \SSA{}
representation such that the (virtual) register pressure does not increase, which
translates to 2\% fewer spills while compiling \SPEC{} benchmark suite.

We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
infrastructure available in \LLVM{}. The experimental results show an average
saving of 1\% on overall compile time on \SPEC{}. GCM enables more inlining and
exposes more loop invariant code motion opportunities in majority of the
benchmarks. We have also seen execution time improvements in a few of \SPEC{}
benchmarks viz. mcf and sjeng, moreover, register spills reduced by 2\% on the
\SPEC{} bencmarks suite when compiled for X86\_64. This is an optimistic
algorithm in the sense that it considers all identical computations in a
function as potential candidates. We make an extra effort to hoist candidates by
partitioning the potential candidates in a way to enable partial hoisting in
case common hoisting points for all the candidates cannot be found. We also
formalize how register pressure will reduce as a result of code-motion, and how
code-motion increases instruction level parallelism thereby enabling more out of
order execution on modern architectures.
\end{abstract}

@@

\subsubsection{Barriers}
Barriers are based on the concept of pinned instructions \cite{click1995global}
but extended to adapt to \LLVM{} IR. Since a basic block in \LLVM{} IR is
actually an extended basic block because there might be calls in the middle of a
basic block which might not return. Essentially, any instruction that cannot
guarantee progress is marked as a barrier. In the absence of context, as in our
current implementation, some instructions which might be safe are still
classified as barriers, e.g., volatile loads/stores, calls with missing
attributes.

\subsubsection{Downward-safety of movable instructions}

In order to be able to move an instruction that has side effects, we must
guarantee that the side-effects of that instructions before and after its motion
will produce the same side-effects.  For hoisting in particular, we need to
prove that on all execution paths from the hoisting point to the end of the
program the side effects appear in the same order.

\begin{verbatim}
computing downward-safety of hoistable instructions at hoist-point
  Worklist = list of all basic blocks of hoistable instructions
  for each basic block B in the dominator tree starting at hoist-point:
    if Worklist is empty
      return false // Path exists! not downward safe
    if B is in Worklist:
      remove B from worklist
      remove subtree with root at B // Available at B => downward safe from B
    if B is a leaf basic block:
      return false  // Path exists! not downward safe
    if B dominates hoist-point:
      return false  // Back edge
  return true
\end{verbatim}

@@

\begin{verbatim}
B0: b = m
    c = n
    if c is true then goto B3 else goto B4

@@

% Explain the overall algorithm, explain how anticipability is computed on demand
\begin{verbatim}
Analyses available: Dominator Tree, DFS Numbering, Memory SSA,

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{|l|r|r|r|r|r|r|r|r|}
      \hline
  &  \multicolumn{2}{|c|}{Hoisted by LICM} &\multicolumn{2}{|c|}{Sunk by LICM} &\multicolumn{2}{|c|}{Functions Inlined} &  \multicolumn{2}{|c|}{Functions Deleted} \\\hline
 Benchmark        & trunk       & GCM	& trunk	& GCM	& trunk	& GCM	& trunk	& GCM  \\\hline
 400.perlbench	  & 2428	& 2480	& 36	& 35	& 1562	& 1565	& 232	& 234  \\\hline
 401.bzip2	  & 721	        & 825	& 3	& 3	& 172	& 180	& 32	& 33   \\\hline
 403.gcc	  & 8867	& 10079	& 143	& 182	& 5034	& 5080	& 1303	& 130  \\\hline
 429.mcf	  & 55	        & 51	& 1	& 1	& 4	& 4	& 0 	& 0    \\\hline
 433.milc	  & 2537	& 2795	& 3	& 3	& 75	& 81	& 0	& 0    \\\hline
 444.namd	  & 2204	& 2206	& 240	& 240	& 387	& 390	& 57	& 55   \\\hline
 445.gobmk	  & 3090	& 3107	& 18	& 19	& 735	& 775	& 230	& 234  \\\hline
 447.dealII	  & 59114	& 95884	& 60	& 8	& 100891& 100692& 30219	& 29966\\\hline
 450.soplex	  & 4668	& 4848	& 18	& 17	& 9801	& 9784	& 3125	& 3112 \\\hline
 453.povray	  & 9533	& 9079	& 23	& 32	& 5522	& 5532	& 1202	& 1201 \\\hline
 456.hmmer	  & 3200	& 3232	& 3	& 8	& 562	& 562	& 103	& 101  \\\hline
 458.sjeng	  & 400	        & 446	& 0	& 0	& 127	& 135	& 18	& 18   \\\hline
 462.libquantum	  & 362	        & 386	& 0 	& 0	& 87	& 87	& 21	& 21   \\\hline
 464.h264ref	  & 17075	& 17153	& 4	& 4	& 338	& 382	& 72	& 72   \\\hline
 470.lbm	  & 39	        & 39	& 0	& 0	& 13	& 13	& 4	& 4    \\\hline
 471.omnetpp	  & 565	        & 661	& 32	& 30	& 4790	& 4843	& 1321	& 1316 \\\hline
 473.astar	  & 691	        & 754	& 0	& 0	& 359	& 363	& 113	& 113  \\\hline
 482.sphinx3	  & 1599	& 1693	& 11	& 9	& 198	& 194	& 53	& 53   \\\hline
 483.xalancbmk	  & 9997	& 11453	& 56	& 54	& 69534	& 68839	& 29875	& 29519\\\hline
    \end{tabular}
  \end{center}
  \caption{Static metrics before and after code-hoisting on SPEC2006 at -Ofast. GCM improves LICM and inlining}
  \label{tab:stats}
\end{table}
